---
title: "Most damaging climate events across the USA"
subtitle: "From 1950 to 2011"
author: "Ander Elkoroaristizabal"
date: "9/2/2020"
output:
  # pdf_document
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

In this analysis we answer the following two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

We will consider the impact of each event type additively, 
i.e., the impact of each event type will be the sum of all the events inside this type 
We leave the option of considering the average damage for later. 

## Synopsis

The analysis can be summarized as follows:

1. We load the data.
2. We start with the first question. We see that the data needs no preprocessing. 
3. We do two barplots, one of injured people and another one of fatalies in order to answer the question.
4. We start with the second question. We see that the data needs preprocessing. We process the exponents data, and modify accordingly the damage columns.
5. We do two barplots of the property and crop damage by type of event and answer the question.
6. We do another barplot, in which the data is the addition of property and crop damage, in order to give more insight about the next events with worst economic consequences.

## Data processing

We download the dataset from the given 
[link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). 
The explanatory documentation may be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

```{r Data_loading, cache = TRUE}
if (!file.exists("StormData.csv.bz2")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                "StormData.csv.bz2")}
Storm_data = read.csv(bzfile("StormData.csv.bz2"))
```

No data preprocessing is needed in order to answer our questions.

<!-- We next process the dataset.  -->
<!-- First we correct the BGN_DATE, BGN_TIME, END_DATE and END_time formatting,  -->
<!-- and add a new columns displaying date and time together. -->

<!-- ```{r Data_processing} -->
<!-- # Formatting of beggining dates & times -->
<!-- Storm_data$BGN_DATE = sapply(strsplit(Storm_data$BGN_DATE," "),`[`, 1) -->
<!-- Storm_data$BGN_TIME = paste(substr(Storm_data$BGN_TIME,1,2), -->
<!--                             substr(Storm_data$BGN_TIME,3,4), -->
<!--                             sep = ":") -->
<!-- Storm_data$BGN_DATETIME = with(Storm_data, paste(BGN_DATE, BGN_TIME)) -->
<!-- Storm_data$BGN_DATETIME = as.POSIXct(Storm_data$BGN_DATETIME,  -->
<!--                                      format = "%m/%d/%Y %H:%M") -->

<!-- # Formatting of ending dates & times -->
<!-- Storm_data$END_DATE = sapply(strsplit(Storm_data$END_DATE," "),`[`, 1) -->
<!-- Storm_data$END_TIME = substr(Storm_data$END_TIME,1,5) -->
<!-- Storm_data$END_DATETIME = with(Storm_data, paste(END_DATE, END_TIME)) -->
<!-- Storm_data$END_DATETIME = as.POSIXct(Storm_data$END_DATETIME,  -->
<!--                                      format = "%m/%d/%Y %H:%M") -->
<!-- ``` -->

## Results

### 1. Most harmful event types (with respect to population health)

To observe the harm done by each event type we will do two barplots, 
one of injured people and another one of fatalies. 
The events shown in the plots are the ones in both the top five injuring and the top five killing.

```{r Harmful, fig.align="center", fig.height = 3, fig.width = 5.5, message = FALSE}
library(dplyr)

hdamage_by_type = group_by(Storm_data, EVTYPE) %>% 
  summarise(INJURIES = sum(INJURIES), FATALITIES = sum(FATALITIES))

most_harmful_events = intersect(hdamage_by_type[order(-hdamage_by_type$INJURIES),][1:5,],
                   hdamage_by_type[order(-hdamage_by_type$FATALITIES),][1:5,])

par(mfrow = c(2,1), mai = c(0.5, 1, 0.1, 0.1), oma = c(0,0,1,0))
barplot(INJURIES/1000~EVTYPE, 
        most_harmful_events,
        xlab = "", ylab = "Injuries (K)", ylim = c(0,100))
title("Human harm by type of event", outer = TRUE)
barplot(FATALITIES/1000~EVTYPE, 
        most_harmful_events,
        xlab = "", ylab = "Fatalities (K)", ylim =c(0,6))
```

The answer to question one is clear: 
Tornados are by far the most harmful event in the US. 

### 2. Event type with worst economic consequences

In order to answer this question we need to focus on the columns PROPDMG and CROPDMG, 
related to property and crops damage, 
and to their exponents, 
in the PROPDMGEXP and CROPDMGEXP columns.
The first thing we notice is that the meaning of some values of these exponent columns is not clear. 
The values are of PROPDMGEXP, for example, are the following:

```{r Exponents}
table(Storm_data$PROPDMGEXP)
```

Most values are understandable, 
but the meaning of the - and + sign is not clear, 
and the question mark should be treated as NA's. 
Nevertheless, since the number of - and + marks is small, 
we will also consider them to be NA's. 
Furthermore, we will treat the empty values as zeros. 

Now we will format the exponent columns using the `translate_exponent()` function,
and use them to give the same scale to the values in PROPDMG and CROPDMG.

```{r Correct_exponents, cache=TRUE}

translate_exponent = function(e){
  if (e==""){ 
    return(0) }
  else if (e %in% c("-","+","?")){
    return(NA)}
  else if (suppressWarnings(!is.na(as.numeric(e)))){ # Check whether it is a number
    return(as.numeric(e))}
  else if (e == "B"){ 
    return(9) } # American billions!
  else if (e %in% c("m","M")){
    return(6) }
  else if (e %in% c("k","K")){
    return(3) }
  else if (e %in% c("h", "H")){
    return(2)}
}

Storm_data$PROPDMGEXP = unlist(sapply(Storm_data$PROPDMGEXP, translate_exponent))
Storm_data$CROPDMGEXP = unlist(sapply(Storm_data$CROPDMGEXP, translate_exponent))
Storm_data = transform(Storm_data, PROPDMG = PROPDMG * 10^(PROPDMGEXP),
                      CROPDMG = CROPDMG * 10^(CROPDMGEXP))
```

Now that the values of CROPDMG and PROPDMG are coherent, 
we proceed as with question 1 and compare the crop and property damages:

```{r Damage_plots, fig.align="center", fig.height = 3.5, fig.width = 10, message = FALSE}

edamage_by_type = group_by(Storm_data, EVTYPE) %>% 
  summarise(PROPDMG = sum(PROPDMG), CROPDMG = sum(CROPDMG))

worst_economic_events = intersect(edamage_by_type[order(-edamage_by_type$PROPDMG),][1:10,],
                   edamage_by_type[order(-edamage_by_type$CROPDMG),][1:10,])

par(mfrow = c(2,1), mai = c(0.5, 1, 0.25, 0.25), oma = c(0,0,1,0))
barplot(PROPDMG/10^9~EVTYPE, 
        worst_economic_events,
        xlab = "", ylab = "Property damage (B$)")
title("Economic damage by event and damage type", outer = TRUE)
barplot(CROPDMG/10^9~EVTYPE, 
        worst_economic_events,
        xlab = "", ylab = "Crop damage (B$)")
```

From the plots above we can see that the floods are the most damaging events in the US, 
both for property and crop. 
Nevertheless, notice that the maximum economic impact due to property damage is much bigger than the maximum economic impact due to crop damage. 
Hence it is sensible to consider it when comparing, 
and see which event is the most harmful when adding the property and crop damage:

```{r Sum_damages, fig.align="center",fig.height = 4, fig.width = 8, message = FALSE}
library(ggplot2)
Pdamage_by_type = group_by(Storm_data, EVTYPE) %>% 
  summarise(Damage = sum(PROPDMG))
Pdamage_by_type["DMG_type"] = "Property"
Cdamage_by_type = group_by(Storm_data, EVTYPE) %>% 
  summarise(Damage = sum(CROPDMG))
Cdamage_by_type["DMG_type"] = "Crop"

# We now take top 3 types of each
Top_Pdamage_types = Pdamage_by_type[order(-Pdamage_by_type$Damage),]$EVTYPE[1:3]
Top_Cdamage_types = Cdamage_by_type[order(-Cdamage_by_type$Damage),]$EVTYPE[1:3]

Edamage_by_type = data.frame(row.names = c("EVTYPE","Damage","DMG_type"))

# And bind the data pairs of the events above
for (type in unique(c(Top_Pdamage_types,Top_Cdamage_types))){
  Edamage_by_type = rbind(Edamage_by_type, Pdamage_by_type[Pdamage_by_type$EVTYPE == type,])
  Edamage_by_type = rbind(Edamage_by_type, Cdamage_by_type[Cdamage_by_type$EVTYPE == type,])
}

ggplot(Edamage_by_type, aes(fill=DMG_type, y=Damage/10^9, x=reorder(EVTYPE,-Damage))) + 
    geom_bar(position="stack", stat="identity") + 
    ggtitle("Events with worst economic consequences") + 
    xlab("") + ylab("Economic damage (B$)") + theme(plot.title = element_text(hjust = 0.5))
```

We can see in the plot above that as we already knew, 
floods are the events that do most economic damage. 
In addition, 
we now see (without any doubt) which are the next 4 most damaging events.